---
title: "2026_AST_MDR_STUDY"
author: "HM"
date: "2026-03-02"
output: html_document
---

```{r, message= F, warning= F}
library(ggplot2)
library(dplyr)
library(tidyr)
library(readr)
library(readxl)

raw <- read_xlsx("../data/2026_ast_data_amik.xlsx", na = c("", "NA"))

head(raw)

```

```{r, message= F, warning= F}
# Clean column names
colnames(raw) <- trimws(colnames(raw))

cat("✔ File loaded:", nrow(raw), "rows x", ncol(raw), "columns\n")
cat("  Columns:", paste(colnames(raw), collapse = ", "), "\n\n")

```

```{r, message= F, warning= F}

# Identify antibiotic columns (everything except ID and Bacteria)
id_col  <- "Isolate_ID"
org_col <- "Bacteria"
ab_cols <- setdiff(colnames(raw), c(id_col, org_col))

ast_long <- raw %>%
  rename(Isolate_ID = all_of(id_col),
         Organism   = all_of(org_col)) %>%
  
  # pivot_longer
  pivot_longer(
    cols      = all_of(ab_cols),
    names_to  = "Antibiotic",
    values_to = "Result"
  ) %>%
  
  filter(!is.na(Result)) %>%                          # drop untested (blank cells)
  mutate(
    Result     = toupper(trimws(Result)),
    Organism   = trimws(Organism),
    Antibiotic = trimws(Antibiotic)
  ) %>%
  
  filter(Result %in% c("S", "I", "R")) %>%            # keep only valid SIR values
  mutate(Result = factor(Result, levels = c("S","I","R"), ordered = TRUE))

cat("✔ Data reshaped:", nrow(ast_long), "observations (blanks excluded)\n")
cat("  Organisms  :", paste(sort(unique(ast_long$Organism)), collapse = " | "), "\n")
cat("  Antibiotics:", paste(sort(unique(ast_long$Antibiotic)), collapse = ", "), "\n\n")


```
# Antibiotics tested per organism
```{r}

tested_matrix <- ast_long %>%
  distinct(Organism, Antibiotic) %>%
  mutate(Tested = "✔") %>%
  pivot_wider(names_from = Antibiotic, values_from = Tested, values_fill = "—")

cat("  Antibiotics tested per organism:\n")
print(as.data.frame(tested_matrix), row.names = FALSE)
cat("\n")

```

#Resistance summary table
```{r}
cat(strrep("=", 65), "\n")
cat("RESISTANCE PROFILES PER ORGANISM\n")
cat(strrep("=", 65), "\n")

summary_table <- ast_long %>%
  group_by(Organism, Antibiotic) %>%
  summarise(
    n           = n(),
    S_n         = sum(Result == "S"),
    I_n         = sum(Result == "I"),
    R_n         = sum(Result == "R"),
    S_pct       = round(S_n / n * 100, 1),
    I_pct       = round(I_n / n * 100, 1),
    R_pct       = round(R_n / n * 100, 1),
    NonSusc_pct = round((I_n + R_n) / n * 100, 1),
    .groups = "drop"
  )

for (org in sort(unique(summary_table$Organism))) {
  sub <- summary_table %>% filter(Organism == org)
  n_isolates <- ast_long %>% filter(Organism == org) %>%
    distinct(Isolate_ID) %>% nrow()
  cat(sprintf("\n  Organism: %-20s | n = %d isolates\n", org, n_isolates))
  cat("  ", strrep("-", 62), "\n", sep = "")
  cat(sprintf("  %-18s %5s %7s  %5s %7s  %5s %7s  %9s\n",
              "Antibiotic","S","S%","I","I%","R","R%","NonSusc%"))
  cat("  ", strrep("-", 62), "\n", sep = "")
  for (i in seq_len(nrow(sub))) {
    cat(sprintf("  %-18s %5d %6.1f%%  %5d %6.1f%%  %5d %6.1f%%  %8.1f%%\n",
                sub$Antibiotic[i],
                sub$S_n[i], sub$S_pct[i],
                sub$I_n[i], sub$I_pct[i],
                sub$R_n[i], sub$R_pct[i],
                sub$NonSusc_pct[i]))
  }
}

```


**MDR ASSESSMENT (Resistant to >= 3 antibiotics)**

```{r}
cat("\n\n", strrep("=", 65), "\n", sep = "")
cat("  MULTI-DRUG RESISTANCE (MDR)\n")
cat("  Resistant (R) to >= 3 antibiotics\n")
cat(strrep("=", 65), "\n\n")

mdr_table <- ast_long %>%
  group_by(Organism, Isolate_ID) %>%
  summarise(R_count = sum(Result == "R"), .groups = "drop") %>%
  group_by(Organism) %>%
  summarise(
    n_isolates = n(),
    MDR_n      = sum(R_count >= 3),
    MDR_pct    = round(MDR_n / n_isolates * 100, 1),
    .groups    = "drop"
  )

print(mdr_table, row.names = FALSE)

```

# Statistical tests 
##Fisher's Exact test / Chi-Square test
Fisher's Exact was done if any expected count was < 5
Chi-Square test, if expected counts >= 5
This only compares organisms that share the SAME antibiotic

```{r}
cat("\n\n", strrep("=", 65), "\n", sep = "")
cat("  INTER-ORGANISM STATISTICAL COMPARISON\n")
cat("  Rule: Fisher's Exact if any expected count < 5\n")
cat("        Chi-Square if all expected counts >= 5\n")
cat("  Only antibiotics tested in BOTH organisms are compared\n")
cat(strrep("=", 65), "\n")

organisms <- sort(unique(ast_long$Organism))
pairs     <- combn(organisms, 2, simplify = FALSE)

all_test_results <- list()

for (pair in pairs) {
  org1 <- pair[1]; org2 <- pair[2]
  d1   <- ast_long %>% filter(Organism == org1)
  d2   <- ast_long %>% filter(Organism == org2)
  shared_abs <- intersect(unique(d1$Antibiotic), unique(d2$Antibiotic))

  if (length(shared_abs) == 0) {
    cat(sprintf("\n  %s vs %s: No shared antibiotics — skipped\n", org1, org2))
    next
  }

  cat(sprintf("\n  ┌── %s  vs  %s\n", org1, org2))
  cat(sprintf("  %-20s %-15s %-12s %-10s %s\n",
              "Antibiotic","Test","Statistic","p-value","Sig."))
  cat("  ", strrep("-", 63), "\n", sep = "")

  for (ab in shared_abs) {
    v1  <- factor(d1$Result[d1$Antibiotic == ab], levels = c("S","I","R"))
    v2  <- factor(d2$Result[d2$Antibiotic == ab], levels = c("S","I","R"))
    tab <- rbind(table(v1), table(v2))
    tab <- tab[, colSums(tab) > 0, drop = FALSE]

    if (ncol(tab) < 2) {
      cat(sprintf("  %-20s %-15s %-12s %-10s\n", ab, "N/A", "—", "uniform"))
      next
    }

    expected   <- outer(rowSums(tab), colSums(tab)) / sum(tab)
    use_fisher <- any(expected < 5)

    if (use_fisher) {
      res       <- fisher.test(tab, simulate.p.value = TRUE, B = 10000)
      test_name <- "Fisher's Exact"
      stat_str  <- "—"
      p_val     <- res$p.value
    } else {
      res       <- chisq.test(tab)
      test_name <- "Chi-Square"
      stat_str  <- paste0("X2=", round(res$statistic, 2))
      p_val     <- res$p.value
    }

    sig   <- ifelse(p_val < 0.001, "***",
             ifelse(p_val < 0.01,  "**",
             ifelse(p_val < 0.05,  "*", "ns")))
    p_str <- ifelse(p_val < 0.001, "<0.001", as.character(round(p_val, 3)))

    cat(sprintf("  %-20s %-15s %-12s %-10s %s\n",
                ab, test_name, stat_str, p_str, sig))

    all_test_results[[paste(org1, org2, ab, sep = "|")]] <-
      data.frame(Org1 = org1, Org2 = org2, Antibiotic = ab,
                 Test = test_name, p_value = p_val, Sig = sig)
  }
}


```


##Kruskal-Wallis test (Within each organism, across antibiotics)
```{r}

cat("\n\n", strrep("=", 65), "\n", sep = "")
cat("  KRUSKAL-WALLIS TEST (within-organism)\n")
cat("  Ordinal scores: S=0, I=1, R=2\n")
cat("  Tests whether resistance levels differ across antibiotics\n")
cat(strrep("=", 65), "\n\n")

ast_long <- ast_long %>%
  mutate(Score = as.integer(Result) - 1)   # S=0, I=1, R=2

for (org in sort(unique(ast_long$Organism))) {
  sub     <- ast_long %>% filter(Organism == org)
  ab_list <- unique(sub$Antibiotic)
  groups  <- lapply(ab_list, function(ab) sub$Score[sub$Antibiotic == ab])
  vary    <- sapply(groups, function(g) length(unique(g)) > 1)
  groups  <- groups[vary]
  labels  <- ab_list[vary]

  if (length(groups) < 2) {
    cat(sprintf("  %-22s: Not enough variation across antibiotics\n", org))
    next
  }

  res   <- kruskal.test(groups)
  sig   <- ifelse(res$p.value < 0.001, "***",
           ifelse(res$p.value < 0.01,  "**",
           ifelse(res$p.value < 0.05,  "*", "ns")))
  p_str <- ifelse(res$p.value < 0.001, "<0.001",
                  formatC(res$p.value, format = "f", digits = 4))

  cat(sprintf("  %-22s: H = %.2f, df = %d, p = %s %s\n",
              org, res$statistic, res$parameter, p_str, sig))
  cat(sprintf("  %sTested: %s\n\n", strrep(" ", 25), paste(labels, collapse = ", ")))
}

```

#Visualization — Stacked Bar Chart per Organism
```{r}
plot_data <- summary_table %>%
  select(Organism, Antibiotic, S_pct, I_pct, R_pct) %>%
  pivot_longer(cols      = c(S_pct, I_pct, R_pct),
               names_to  = "Category",
               values_to = "Percentage") %>%
  mutate(
    Category = recode(Category, "S_pct" = "S", "I_pct" = "I", "R_pct" = "R"),
    Category = factor(Category, levels = c("R","I","S"))
  )

sir_colors <- c("S" = "#2ecc71", "I" = "#f39c12", "R" = "#e74c3c")

p <- ggplot(plot_data, aes(x = Antibiotic, y = Percentage, fill = Category)) +
  geom_bar(stat = "identity", position = "stack",
           color = "white", linewidth = 0.3) +
  facet_wrap(~Organism, scales = "free_x", ncol = 2) +
  scale_fill_manual(
    values = sir_colors,
    labels = c("S" = "Sensitive", "I" = "Intermediate", "R" = "Resistant"),
    name   = "Susceptibility"
  ) +
  scale_y_continuous(labels = scales::label_percent(scale = 1)) +
  labs(
    title    = "Antimicrobial Susceptibility Profiles by Organism",
    x        = "Antibiotic",
    y        = "Percentage of Isolates (%)"
  ) +
  theme_bw(base_size = 11) +
  theme(
    axis.text.x      = element_text(angle = 40, hjust = 1, size = 9),
    strip.background = element_rect(fill = "#2c3e50"),
    strip.text       = element_text(color = "white", face = "bold", size = 10),
    legend.position  = "bottom",
    plot.title       = element_text(face = "bold", size = 13),
    plot.subtitle    = element_text(size = 10, color = "gray40"),
    panel.grid.minor = element_blank()
  )

print(p)

ggsave("AST_Plot.png", p, width = 13, height = 9, dpi = 300)

```


```{r}

write.csv(summary_table, "AST_Resistance_Summary.csv", row.names = FALSE)
write.csv(mdr_table,     "AST_MDR_Summary.csv",        row.names = FALSE)

if (length(all_test_results) > 0) {
  test_df <- do.call(rbind, all_test_results)
  write.csv(test_df, "AST_Statistical_Tests.csv", row.names = FALSE)
}

cat("\n✔ Exported:\n")
cat("  → AST_Resistance_Summary.csv\n")
cat("  → AST_MDR_Summary.csv\n")
cat("  → AST_Statistical_Tests.csv\n\n")
cat(strrep("=", 65), "\n")
cat("Significance: *** p<0.001 | ** p<0.01 | * p<0.05 | ns = not significant\n")
cat(strrep("=", 65), "\n")

```
